---
title: "A report on lake ontario's microbes"
output: html_document
date: "2025-02-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#this sets global changes
```
#prepare the R environment
```{r packages}
#load library/packages
library(tidyverse)
```

#load in the lake ontario data
```{r}
#load in lake ontario microbial community data
sample_and_taxon<-
 read_csv("data/sample_and_taxon.csv")

#inspect the data
glimpse(sample_and_taxon)
```

#microbioal abundance versus temperature
```{r cells_vs_temp}
#temp on x, abundance on y, colored by env_group

ggplot(data=sample_and_taxon,
       aes(x=temperature, y=cells_per_ml))+
  geom_point(aes(color=env_group))+
  labs(x="Temp(Celsius)",y="Cell Density (cells/mL")+
  theme_classic()+
  scale_color_manual(values = c("cornflowerblue",
                              "goldenrod1",
                              "firebrick"))+
  geom_smooth(method = "lm",color="black")+
  theme(legene.position="bottom",
        legend.title = element_blank())

```

The above plot shows that:

-there's a positive relationship and cell abundances

-for example, deep samples are the coldest and have fewest cells

```{r data summary, include = FALSE}
#how many samples do I have
n_samples<-
  sample_and_taxon%>%
  #only pick out sample ID column
  select(sample_id)%>%
  #number of distinct names in the column
  n_distinct()
  
n_samples

#calcualte the minimum and maximum temps
#min
min_temp<-
sample_and_taxon%>%
  summarize(round(min(temperature), digits=1))%>%
  pull()
#max
max_temp<-
sample_and_taxon%>%
  summarize(round(max(temperature), digits=1))%>%
  pull()

```

The total number of samples is `r n_samples`. For this set of samples, temperature ranged from a minimum of `r min_temp` celsius to a maximum of `r max_temp` celsius.
